Rectangulotopes

### 2024-04-03 draft/rectangulotopes.tex
details

\begin{abstract}
  Rectangulations are decompositions of a square into finitely many axis-aligned rectangles.
  We describe realizations of $(n-1)$-dimensional polytopes associated with rectangulations composed of $n$ rectangles. They are defined as quotientopes of natural congruence relations on the weak Bruhat order on permutations in $\f{S}_n$, and their skeleta are flip graphs on rectangulations. We give a thorough description of these polytopes, including elementary formulas for computing the coordinates of each vertex of the polytope, in the spirit of Loday's realization of the associahedron.
\end{abstract}

@@

The relevant lattice congruences are the \defn{Baxter congruence} on one hand, studied by Law and Reading~\cite{MR2871762}, and whose classes are one-to-one with \defn{weak rectangulations} (or \defn{diagonal rectangulations}) and \defn{Baxter permutations}, and the \defn{strong rectangulation congruence}, whose classes are in bijection with \defn{strong rectangulations}~\cite{MR2864445}. The latter is mentioned by Reading~\cite{MR3335492} (see for instance in Theorem 1.2, and Example 4.11), and studied more recently by Asinowski et al.~\cite{ACFF24}.

One of our main results is an elementary description of the vertex coordinates of a realization of both families of quotientopes, in the spirit of Loday's realization in Theorem~\ref{thm:loday}.
The quotientopes corresponding to the Baxter congruence will be referred to as the \defn{weak rectangulotopes} and denoted by $\WRP(n)$.
The skeleton of $\WRP(n)$ is isomorphic to the flip graph of weak rectangulations of size $n$ defined by Law and Reading~\cite{MR2871762}.
Similarly, the quotientopes corresponding to the strong rectangulation congruence will be called the \defn{strong rectangulotopes} and denoted by $\SRP(n)$.
The skeleton of $\SRP(n)$ is isomorphic to the flip graph of strong rectangulations of size $n$ defined by Meehan~\cite{MR3697823} (see also~\cite{ACFF24}).

@@

%%%%%%%%%%%%
\subsection{The Baxter congruence}

% jean:{Clash of terminology: weak vs. Baxter?}

@@

We call this congruence the \defn{Baxter congruence}.
The \defn{weak rectangulotope} $\WRP(n)$ is the quotientope of the Baxter congruence $\baxtereq$ on the weak Bruhat order on $\f{S}_n$.

@@

\begin{lemma}
The Baxter congruence $\baxtereq$ on $\f{S}_n$ is defined by the arc ideal composed of arcs that do not cross the horizontal line.
\end{lemma}

@@

\begin{lemma}
  The weak rectangulotope $\WRP(n)$ is realized by the Minkowski sum of all Loday and anti-Loday shard polytopes. In particular, it is the Minkowski sum of two opposite Loday associahedra.
\end{lemma}

@@

\begin{lemma}
  The weak rectangulotope $\WRP(n)$ is realized as
  \[
  \{x\in\R^n : \sum_{i\in S} x_i \leq f(S) \text { and } \sum_{i\in [n]} x_i = f([n])\},
  \]
  where $f:2^{[n]}\mapsto \R$ is the submodular function that maps $S$ to the number of intervals
  $I$ of $[n]$ such that $I\cap S\not=\emptyset$ and $I\not\subseteq S$. 
\end{lemma}

@@

Note that this extremal vertex should only depend on the Baxter congruence class of $\pi$.
This, in passing, gives us an alternative characterization of the Baxter congruence:
$\pi\baxtereq\sigma$ if and only if for any interval $I$ of $[n]$,
%the maximum and the minimum of $\pi^{-1}$ and $\sigma^{-1}$ on $I$ have the same index.
$\arg\max_{j\in I} \pi^{-1}(j)=\arg\max_{j\in I} \sigma^{-1}(j)$, and
$\arg\min_{j\in I} \pi^{-1}(j)=\arg\min_{j\in I} \sigma^{-1}(j)$.

@@

% jean:{Does the congruence have a name?}

@@

% jean:{From this, do we directly have a description of the face lattice of $\SRP(n)$?}

@@

% jean:{From this, do we directly have the submodular function defining $\SRP(n)$?}

>>>>

\begin{abstract}
  Rectangulations are decompositions of a square into finitely many axis-aligned rectangles.
  We describe realizations of $(n-1)$-dimensional polytopes associated with rectangulations composed of $n$ rectangles.
  They are defined as quotientopes of natural congruence relations on the weak Bruhat order on permutations in $\f{S}_n$, and their skeleta are flip graphs on rectangulations.
  We give a thorough description of geometric realizations of these polytopes, including elementary formulas for computing the coordinates of each vertex of the polytope, in the spirit of Loday's realization of the associahedron.
\end{abstract}

@@

The relevant lattice congruences are the \defn{weak rectangulation congruence} on one hand, studied by Law and Reading~\cite{MR2871762}, and whose classes are one-to-one with \defn{weak rectangulations} (or \defn{diagonal rectangulations}) and \defn{Baxter permutations}, and the \defn{strong rectangulation congruence}, whose classes are in bijection with \defn{strong rectangulations}~\cite{MR2864445}. The latter is mentioned by Reading~\cite{MR3335492} (see for instance in Theorem 1.2, and Example 4.11), and studied more recently by Asinowski et al.~\cite{ACFF24}.

One of our main results is an elementary description of the vertex coordinates of a realization of both families of quotientopes, in the spirit of Loday's realization in Theorem~\ref{thm:loday}.
The quotientopes corresponding to the weak rectangulation congruence will be referred to as the \defn{weak rectangulotopes} and denoted by $\WRP(n)$.
The skeleton of $\WRP(n)$ is isomorphic to the flip graph of weak rectangulations of size $n$ defined by Law and Reading~\cite{MR2871762}.
Similarly, the quotientopes corresponding to the strong rectangulation congruence will be called the \defn{strong rectangulotopes} and denoted by $\SRP(n)$.
The skeleton of $\SRP(n)$ is isomorphic to the flip graph of strong rectangulations of size $n$ defined by Meehan~\cite{MR3697823} (see also~\cite{ACFF24}).

@@

%%%%%%%%%%%%
\subsection{The weak rectangulation congruence}

@@

We call this congruence the \defn{weak rectangulation congruence}.
The \defn{weak rectangulotope} $\WRP(n)$ is the quotientope of the weak rectangulation congruence $\baxtereq$ on the weak Bruhat order on $\f{S}_n$.

@@

\begin{lemma}
The weak rectangulation congruence $\baxtereq$ on $\f{S}_n$ is defined by the arc ideal composed of arcs that do not cross the horizontal line.
\end{lemma}

@@

\begin{lemma}
  \label{lem:weakMinkowski}
  The weak rectangulotope $\WRP(n)$ is realized by the Minkowski sum of all Loday and anti-Loday shard polytopes. In particular, it is the Minkowski sum of two opposite Loday associahedra.
\end{lemma}

@@

Weak rectangulotopes are deformed permutahedra, hence are defined by a \defn{submodular} set function $f:2^{[n]}\mapsto \R$. The description of the weak rectangulotope $\WRP(n)$ as a Minkowski sum in Lemma~\ref{lem:weakMinkowski} allows us to easily compute the submodular function defining it.

\begin{lemma}
  The weak rectangulotope $\WRP(n)$ is realized as
  \[
  \{x\in\R^n : \sum_{i\in S} x_i \leq f(S) \text { and } \sum_{i\in [n]} x_i = f([n])\},
  \]
  where $f:2^{[n]}\mapsto \R$ is the following submodular function:
  \[
  f(S) := |\{ I\text{ interval of } [n] : I\cap S\not=\emptyset \text{ and } I\cap ([n]\setminus S) \not= \emptyset \}|.
  \]
  %that maps $S$ to the number of intervals $I$ of $[n]$ such that $I\cap S\not=\emptyset$ and $I\not\subseteq S$. 
\end{lemma}

The proof relies on the following simple lemma stating that the submodular function of a Minkowski sum of deformed permutahedra is the sum of the submodular function of the summands.

\begin{lemma}
  \label{lem:submodsum}
  Consider a collection $\polytope{P}_1, \polytope{P}_2,\ldots ,\polytope{P}_k$ of deformed permutahedra defined by the submodular functions $f_1,f_2,\ldots ,f_k$. Then the Minkowski sum $\sum_i \polytope{P}_i$ is defined by the submodular function $f:=\sum_i f_i$.
\end{lemma}

\begin{proof}
  From Lemmas~\ref{lem:weakMinkowski} and~\ref{lem:submodsum}, the submodular function defining the realization of $\WRP(n)$ is the sum of the submodular functions defining the Loday and anti-Loday shard polytope in Lemmas~\ref{lem:lodaysp} and~\ref{lem:antilodaysp}. For a Loday arc $A$ defined by an interval $I$ of $[n]$, we have
  \[
  \SP(A) = \{ x\in\R^n : \sum_{i\in S} x_i \leq \loday{f}_I(S) \text { and } \sum_{i\in [n]} x_i = \loday{f}_I([n]) \},
  \]
  where
  \[
  \loday{f}_I(S) := \llbracket S\cap I\not=\emptyset \rrbracket .
  \]
  Similarly, for an anti-Loday arc $A$ defined by an interval $I$, the shard polytope $\SP(A)$ is defined by the following submodular function:
  \[
  \antiloday{f}_I(S) := - \llbracket S\subseteq I \rrbracket .
  \]
  Their sum
  \[
  f(S) := \sum_{I\text{ interval of }[n]} \loday{f}_I(S) + \antiloday{f}_I(S)
  \]
  is therefore exactly the number of intervals of $I$ that intersect $S$ but are not contained in $S$, hence the number of intervals intersecting both $S$ and its complement.
\end{proof}

@@

Note that this extremal vertex should only depend on the weak rectangulation congruence class of $\pi$.
This, in passing, gives us an alternative characterization of the weak rectangulation congruence:
$\pi\baxtereq\sigma$ if and only if for any interval $I$ of $[n]$,
%the maximum and the minimum of $\pi^{-1}$ and $\sigma^{-1}$ on $I$ have the same index.
$\arg\max_{j\in I} \pi^{-1}(j)=\arg\max_{j\in I} \sigma^{-1}(j)$, and
$\arg\min_{j\in I} \pi^{-1}(j)=\arg\min_{j\in I} \sigma^{-1}(j)$.

@@

@@

@@

\jean{Compute the submodular functions of each yin and yang shard polytopes and sum them.}
